Internal Gravity Wave Breaking in White Dwarf Binaries

### 2019-04-16 writeups/paper.tex
stash paper writeup

\title[IGW Breaking in WDs]{Tidal Heating via Internal Gravity Waves in White Dwarf Binaries}
\author[Y. Su et\ al.]{
Yubo Su,$^1$,
Daniel Lecoanet,$^2$
Dong Lai,$^1$
\\
$^1,^3$ ABC
$^2$ DEF
}

@@

\begin{document}\label{firstpage}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\maketitle

@@

\begin{keywords}
Wave Breaking, Dynamical Tides, Internal Gravity Waves
\end{keywords}

@@

Compact white dwarf (WD) binary systems, with orbital periods in the range of
minutes to hours, are important for a range of astrophysical problems. They are
the most important sources of gravitational waves (GWs) for the Laser
Interferometric Space Antenna (LISA)\cite{lisa}. They are also thought to
produce interesting optical transients such as underluminous
supernovae\cite{underlum}, Ca-rich fast transients\cite{carich}, and tidal
novae\cite{tidal_novae}. Most importantly, they have been proposed as the likely
progenitors of type Ia supernovae (e.g.~\cite{Ia0,webbink} or more
recently\cite{Ia1,Ia2}). While presently only a few tens of compact WD binaries
are known\cite{lsst_wd}, \emph{Gaia} (currently gathering data) is expected to
expand the catalog to a few hundreds\cite{lsst_wd} (results based on
\emph{Gaia}'s second data release have already begun to
appear\cite{gaiaDD,gaiaDD2}), and the Large Synoptic Survey Telescope (LSST,
first light scheduled for 2020) will likely detect a few thousand
more\cite{lsst_wd}. These observations will significantly advance the
understanding of WD binaries and their evolution. My proposed theoretical and
computational research is well-timed to take advantage of these new advances.

In spite of the broad importance of WD binaries, the evolution of these systems
prior to their final mergers is not well understood. Much of this uncertainty
comes from our imprecise understanding of tidal interactions, which play an
important role during a compact WD binary's inspiral\cite{fullerII}. Previous
studies have shown that these interactions manifest as tidal excitation of
internal gravity waves (IGW), waves in the WD fluid restored by the buoyancy
force due to density stratification\cite{fullerI}. As these waves propagate
outwards towards the WD surface, they grow in amplitude until they break, as do
ocean waves on a shore, and transfer both energy and angular momentum from the
binary orbit to the outer envelope of the WD\cite{fullerI,fullerII}.

Previous works have found that the dissipation of IGW can generate significantly
more energy than thermal radiation from the isolated WD surface and is thus a
major contributor to the WD energy budget\cite{fullerII,fullerIV}. However,
these works parameterized the wave breaking process in an ad hoc manner. The
details of dissipation, namely the location and spatial extent of the wave
breaking, affect the observable outcome: dissipation near the surface of the WD
can be efficiently radiated away and simply brightens the WD, while dissipation
deep in the WD envelope causes an energy buildup that results in energetic
flares\cite{tidal_novae}. Works in other fields based on numerical simulations
show that strongly nonlinear wave breaking behaves differently than predictions
based in linear and weakly nonlinear theory\cite{winters1994,barker_ogilvie}.
Such fully nonlinear numerical simulations have not been performed for WDs.

@@

The Euler equations for an incompressible, barotropic fluid in a uniform
gravitational field are
\begin{subequations}\label{se:nl_orig}
    \begin{align}
        \vec{\nabla} \cdot \vec{u} &= 0,\\
        \md{\rho}{t} &= 0 ,\label{eq:nl_density}\\
        \md{\vec{u}}{t} + \frac{\vec{\nabla}p}{\rho} + g\hat{z} &=
            0.\label{eq:nl_mom}.
    \end{align}
\end{subequations}
$\md{}{t} = \pd{}{t} + \p*{\vec{u} \cdot \vec{\nabla}}$ is the Lagrangian or
material derivative. $\vec{u}, \rho, p$ denote the velocity field, density and
pressure respectively. We denote $-g\hat{z}$ constant gravitational
acceleration. Note that at hydrostatic equilibrium $\pd{}{t} = 0$ we have
$\vec{\nabla}P_0 = -\rho_0 g\hat{z}$ and so $P_0 = \rho_0 gH$. A shear flow
$u_x(x, z, t) = u_x(z)$ is permitted in hydrostatic equilibrium, but we will
initially consider no shear flow in equilibrium.

@@

\section{Internal Gravity Wave Theory}\label{s:theory}

@@

In the small perturbation limit, where flow velocities are small compared to the
characteristic space and time scales $\pd{}{t} \gg \vec{u} \cdot \vec{\nabla}$,
we may linearize \autoref{se:nl_upsilon}. We ignore the advective components of the
material derivative and consider small deviations $\Upsilon_1 = \Upsilon -
\Upsilon_0, P_1 = P - P_0$ about hydrostatic equilibrium:
\begin{subequations}\label{se:lin_homo}
    \begin{align}
        \vec{\nabla} \cdot \vec{u} &= 0,\\
        \pd{\Upsilon_1}{t} - \frac{u_{z}}{H} &= 0,\\
        \pd{\vec{u}}{t} + \vec{\nabla}P_1
            + \p*{1 + \Upsilon_1} g\hat{z}
            &= 0.
    \end{align}
\end{subequations}

This can be solved to obtain solutions up to undetermined amplitude $A$ the
well-known result\cite{drazin,sutherland0}:
\begin{subequations}\label{se:lin_sol}
    \begin{align}
        u_z\p*{x, z, t} &= Ae^{z/2H}e^{i(k_{0x}x + k_{0z}z - \omega_0 t)},\\
        \omega_0^2 &= \frac{N^2k_{0x}^2}{k_{0x}^2 + k_{0z}^2 + \frac{1}{4H^2}},
            \label{eq:disp_rel}
    \end{align}
\end{subequations}
where
\begin{equation}
    N^2 \equiv g^2\p*{\rd{\rho}{P} - \at{\pd{\rho}{P}}_{ad}}
        = \frac{g}{H},
\end{equation}
the Brunt-V\"ais\"al\"a frequency is constant

@@

    \item The phase and group velocities can be computed respectively:
        \begin{align}
            \vec{c}_{ph} &=
                \p*{\frac{\hat{x}}{k_{0x}} + \frac{\hat{z}}{k_{0z}}}\omega,\\
            \vec{c}_{g} &= \frac{N\p*{k_{0x}k_{0z}\hat{x}
                - \p*{k_{0z}^2 + \frac{1}{4H^2}}\hat{z}}}
                {\p*{k_{0x}^2 + k_{0z}^2 + \frac{1}{4H^2}}^{3/2}} \nonumber\\
                &\approx \frac{\omega k_{0z}}{\abs*{\vec{k}_0}^{3/2}}
                    \p*{k_{0x}\hat{x} - k_{0z}\hat{z}}.
        \end{align}
        We recover the usual result $\vec{c}_{ph} \cdot \vec{c}_g = 0$
        (e.g.\ \cite{drazin,sutherland1}). Note that an IGW transporting energy
        and momentum upwards $c_{g, z} > 0$ has $k_{0z} < 0$.

    \item The time-averaged total $x$-momentum flux in the $\hat{z}$ direction
        can be computed. Since the linear solution is separable as $f(x, z, t) =
        f(z)e^{i(k_{0x}x - \omega_0 t)}$, $x$ averaging and time averaging are
        equivalent, so we may write
        \begin{align}
            S_{px, z} &\equiv \ev*{\rho u_x u_z}_x \equiv
                \frac{1}{L_x}\int_0^{L_x}\limits \rho u_x u_z\;\mathrm{d}x
                    \nonumber\\
                &\approx -\frac{A^2}{2}\rho_0(z=0)\frac{k_{0z}}{k_{0x}},
                    \label{eq:fpx_lin}
        \end{align}
        where $\ev*{\dots}_x$ denotes $x$-averaging.
\end{itemize}

@@

As not to interfere with the incompressibility constraint, we force the system
on the density equation. This constitutes replacing \autoref{eq:nl_up_density}
with
\begin{equation}
    \md{\Upsilon}{t} + u_z\pd{\ln \rho_0}{z}
        = Fe^{-\frac{(z - z_0)^2}{2\sigma^2}}
            \cos \p*{k_{0x}x - \omega_0 t}.\label{eq:vol_drive}
\end{equation}
Using a narrow Gaussian profile excites a broad $k_z$ wavenumber spectrum, and
only the $k_{0z}$ satisfying dispersion relation \autoref{eq:disp_rel} for the
given $k_{0x}, \omega_0(k_{0x}, k_{0z})$ will propagate.

In the linearized system, the effect of this forcing can be solved analytically
up to good accuracy: we first approximate the driving term using
$e^{\frac{-(z - z_0)^2}{2\sigma^2}} \approx \sqrt{2\pi \sigma^2}\delta(z -
z_0)$, the $\sigma \to 0$ limit\footnote{In practice, $\sigma$ must be large
enough to be numerically resolved by the spectral code.}. This system is solved
exactly by matching the two homogeneous solutions above and below $z_0$. We may
then approximately relax the solution to nonzero $\sigma$: an extra factor of
$e^{-\frac{(k_{0z}\sigma)^2}{2}}$ arises compared to the $\delta$-function
solution (evaluating the Fourier Transform of $e^{-\frac{(z -
z_0)^2}{2\sigma^2}}$ at $k_z = k_{0z}$), and we obtain
\begin{align}
    u_{1z}&(x, z, t) ={} \frac{Fgk_{0x}^2}{\omega_0^2}
        \frac{1}{2ik_{0z}}\frac{e^{\frac{-(k_{0z}\sigma)^2}{2}}}
        {\sqrt{2\pi\sigma^2}} \times\nonumber\\
        &{}\begin{cases}
        e^{\frac{z - z_0}{2H}}e^{i\p*{k_{0x}x + k_{0z}(z - z_0) - \omega_0 t
            + \frac{1}{2k_{0z}H}}}
            & z > z_0\\
        e^{\frac{z - z_0}{2H}}e^{i\p*{k_{0x}x - k_{0z}(z - z_0) - \omega_0 t
            + \frac{1}{2k_{0z}H}}}
            & z < z_0\\
    \end{cases}.\label{eq:uz_lin}
\end{align}
The $z > z_0$ solution models an upwards-propagating IGW wavetrain inbound on
the simulation domain from below. The solution for $u_{1x}(x, z, t)$ is
correspondingly $u_{1x} \approx \frac{k_{0z}
u_{1z}}{k_{0x}}e^{\frac{i}{2k_{0z}H}}$.

@@

In developing the results above, we have neglected advective terms $\pd{}{t} \gg
\vec{u} \cdot \vec{\nabla}$. However, since $\abs*{\vec{u}} \propto e^{z/2H}$,
any nonzero amplitude IGW will eventually violate the linearity criterion. This
regime is relevant in practice since the IGW may grow to nonlinear amplitudes
before it reaches the surface of the WD\cite{fullerI,fullerII}. We can coarsely
estimate the height at which this happens by when the Lagrangian displacement
$\vec{\xi}$ of a fluid parcel satisfies
\begin{equation}
    \vec{\xi} \cdot \vec{k} \gtrsim 1.\label{eq:nl}
\end{equation}

The general understanding of nonlinear IGW breaking is that the waves'
interactions with the mean flow of the fluid transfer horizontal momentum from
the waves into the fluid; such a process has been conjectured to be responsible
for tidal synchronization in stellar binaries\cite{zahn75,gn89} as well as the
quasi-biennial oscillation\cite{lindzen_qbo}. The details of this process have
been laid out in a few key papers: IGWs are globally unstable to resonant
three-wave interactions\cite{drazin}, causing energy transfer out of the IGW to
daughter modes. These steeper daughter modes facilitate wave breaking,
depositing horizontal momentum in the fluid's mean flow\cite{klostermeyer}. Once
the mean flow reaches the horizontal phase velocity of the parent IGW, a
critical layer forms at which the frequency of the IGW is Doppler-shifted to
zero. The interaction of the parent IGW with the mean flow was first studied in
the inviscid, linear regime in~\cite{booker_bretherton}, which found nearly
complete absorption. This result was reproduced with nonzero
viscosity\cite{hazel}, but weakly nonlinear theory\cite{brown_stewartson} and
fully nonlinear simulations\cite{winters1994} suggest that nonlinear effects can
induce reflection.

An estimate of where these mean flow effects become important can be made using
the wave-induced mean flow\cite{eliassen_palm_cite,sutherland0}:
\begin{equation}
     \ev*{u_x} = \bar{U}_0 \equiv \frac{1}{L_x}
        \int\limits u_x\;\mathrm{d}x = \frac{\ev*{u_xu_z}_x}{c_{g,z}},
\end{equation}
where $\ev*{\dots}_x$ denotes horizontal averaging. Setting $\bar{U}_0 = c_{ph,
x}$ the critical layer condition, we recover $\frac{u_xk_x}{\omega} = \xi_x k_x
\sim 1$, roughly equivalent to the nonlinearity criterion \autoref{eq:nl}.

\section{Internal Gravity Wave Numerical Simulation}\label{s:sim}

To study IGW breaking as a weak perturbation grows to nonlinear amplitudes, we
perform direct numerical simulation using Dedalus\cite{dedalus} a spectral
numerical method. Our simulation domain is smaller than our full physical
domain, so we adopt reflection-suppressing boundary conditions. We use periodic
boundary conditions in the $x$ direction and damping layers (described in
\autoref{ss:damping}) in the $z$ direction to damp perturbations exiting the
top and bottom of the domain. We use a Fourier basis in the $x$ direction and
either a Chebyshev or Fourier basis in the $z$ direction; our results were
robust to either set of basis functions. We varied the number of $x, z$ modes
(denoted $N_x, N_z$ respectively), and used $3/2$ dealiasing\cite{boyd}.

The time integration uses a split implicit-explicit first-order scheme where
certain terms are treated implicitly using for- ward Euler, and the remaining
terms are treated explicitly using backward Euler. A third-order, four-stage
DIRK-ERK scheme\cite{ascher} is used with adaptive timesteps taken to be
the advective Courant-Friedrichs-Lewy (CFL) time given by $0.7 \min (\Delta
x / u_x,\Delta z / u_z)$. The minimum is taken over every grid point in
the domain and $\Delta x,\Delta z$ are the grid spacings in the $x$ and $z$
directions respectively.

\subsection{Numerical Parameters}\label{ss:numerics}

@@

Denote $L_x, L_z$ to be the physical dimensions of the simulation domain. The
physics of our simulation is fixed by six parameters: $L_x, L_z$, and $k_{0x},
\omega_0, F, \nu$. We describe and choose these parameters below:
\begin{itemize}
    \item $L_z$: Since the wave amplitude grows $\sim e^{z/2H}$, we choose $L_z
        = 10H$ to give $\sim e^4$ apmlitude growth between the damping zones.

@@

    \item $\omega_0$: We choose $\omega_0$ by evaluating dispersion relation
        $\omega_0\p*{k_{0x}, k_{0z}}$ for some desired $k_{0z}$.

@@

    \item $F$: We first choose $F$ forcing strength such that $\vec{\xi} \cdot
        \vec{k} \ll 1$ is satisfied everywhere throughout the domain. This
        becomes a constraint on $F$ by using \autoref{eq:uz_lin} and
        $\abs*{\vec{u}} \propto e^{z/2H}$.

    \item $\nu$: Nonlinear effects transfer wave energy from $\vec{k}_{0}$ to
        smaller scales. Since well-resolved simulations using spectral methods
        have no inherent numerical viscosity, energy will accumulate at grid
        scales in the absence of artificial dissipation. We introduce
        Navier-Stokes viscosity $\nu$ while ensuring our equations of motion
        continue to conserve horizontal momentum (see \autoref{ss:strat_impl}
        for details). When $\vec{\xi} \cdot \vec{k} \ll 1$, we may set $\nu =
        0$.
\end{itemize}

@@

To imitate an infinite fluid using a finite simulation domain, we use periodic
boundary conditions in the $x$ direction and damping layers at the top and
bottom of the $z$ direction. This damps waves that reach the edge of the
simulation domain without inducing nonphysical reflection. Specifically, these
are implemented by introducing linear damping terms to \autoref{se:nl_upsilon}:
\begin{align}
    \md{\Upsilon}{t} + u_z \pd{\ln \rho_0}{z} &=
        -\Gamma(z)\Upsilon,\nonumber\\
    \md{\vec{u}}{t} + \vec{\nabla}P + P\vec{\nabla}\Upsilon + g\hat{z} &=
        - \Gamma(z)\vec{u},\nonumber\\
    \frac{1}{2\tau}\s*{2 + \tanh \frac{z - z_T}{\Delta z}
        + \tanh \frac{z_B - z}{\Delta z}} &= \Gamma(z),\label{eq:Gamma}
\end{align}
where $z_B = 0.05L_z, z_T = 0.95L_z$ are the boundaries of the damping zone.
This strongly damps perturbations below $z_B$ and above $z_T$ with damping time
$\tau$, negligibly affects dynamics between $z_B, z_T$ and has transition width
governed by $\Delta z$; we use $\Delta z = 0.025L_z$. This prescription, used in
similar studies\cite{lecoanet_damp}, has the advantage of being smooth,
important for spectral methods. The full implementation of the fluid equations
in Dedalus is described in \autoref{ss:strat_impl}.

\subsection{Linear Regime}\label{ss:lin_ns}

We first test our numerical simulation with a ``linear'' simulation, where
$F$ is chosen such that $\vec{\xi} \cdot \vec{k} \ll 1$ everywhere. We emphasize
that our ``linear'' simulations solve the full nonlinear equations, just in the
weak perturbation limit.

@@

>>>>

\title[Internal Gravity Wave Breaking in White Dwarf Binaries]{Internal Gravity Wave Breaking in White Dwarf Binaries}
\author[Y. Su et\ al.]{
Yubo Su,$^1$,
Daniel Lecoanet,$^2$
Dong Lai,$^1$
\\
$^1$ Cornell Center for Astrophysics and Planetary Science, Department of
Astronomy, Cornell University, Ithaca, NY 14853, USA
\\
$^2$ Princeton Center for Theoretical Science, Princeton University, Princeton,
NJ 08544, USA
}

@@

\begin{document}\label{firstpage}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\renewcommand*{\sectionautorefname}{Section}
\maketitle

@@

\begin{keywords}
white dwarfs -- hydrodynamics -- binaries:close -- waves % chktex 8
\end{keywords}

@@

Compact white dwarf (WD) binary systems, with orbital periods in the range of
minutes to hours, are important for a range of astrophysical problems. They are
the most important sources of gravitational waves (GWs) for the Laser
Interferometric Space Antenna (LISA) \citep{lisa}. They are also thought to
produce interesting optical transients such as underluminous
supernovae \citep{underlum}, Ca-rich fast transients \citep{carich}, and tidal
novae \citep{tidal_novae}. Most importantly, they have been proposed as the likely
progenitors of type Ia supernovae (e.g. \citep{Ia0,webbink} or more
recently \citep{Ia1,Ia2}). While presently only a few tens of compact WD binaries
are known \citep{lsst_wd}, \emph{Gaia} (currently gathering data) is expected to
expand the catalog to a few hundreds \citep{lsst_wd} (results based on
\emph{Gaia}'s second data release have already begun to
appear \citep{gaiaDD,gaiaDD2}), and the Large Synoptic Survey Telescope (LSST,
first light scheduled for 2020) will likely detect a few thousand
more \citep{lsst_wd}. These observations will significantly advance the
understanding of WD binaries and their evolution. My proposed theoretical and
computational research is well-timed to take advantage of these new advances.

In spite of the broad importance of WD binaries, the evolution of these systems
prior to their final mergers is not well understood. Much of this uncertainty
comes from our imprecise understanding of tidal interactions, which play an
important role during a compact WD binary's inspiral \citep{fullerII}. Previous
studies have shown that these interactions manifest as tidal excitation of
internal gravity waves (IGW), waves in the WD fluid restored by the buoyancy
force due to density stratification \citep{fullerI}. As these waves propagate
outwards towards the WD surface, they grow in amplitude until they break, as do
ocean waves on a shore, and transfer both energy and angular momentum from the
binary orbit to the outer envelope of the WD \citep{fullerI,fullerII}.

Previous works have found that the dissipation of IGW can generate significantly
more energy than thermal radiation from the isolated WD surface and is thus a
major contributor to the WD energy budget \citep{fullerII,fullerIV}. However,
these works parameterized the wave breaking process in an ad hoc manner. The
details of dissipation, namely the location and spatial extent of the wave
breaking, affect the observable outcome: dissipation near the surface of the WD
can be efficiently radiated away and simply brightens the WD, while dissipation
deep in the WD envelope causes an energy buildup that results in energetic
flares \citep{tidal_novae}. Works in other fields based on numerical simulations
show that strongly nonlinear wave breaking behaves differently than predictions
based in linear and weakly nonlinear theory \citep{winters1994,barker_ogilvie}.
Such fully nonlinear numerical simulations have not been performed for WDs.

@@

The Euler equations for an incompressible, barotropic fluid in a uniform
gravitational field are
\begin{subequations}\label{se:nl_orig}
    \begin{align}
        \vec{\nabla} \cdot \vec{u} &= 0,\label{eq:nl_incomp}\\
        \md{\rho}{t} &= 0 ,\label{eq:nl_density}\\
        \md{\vec{u}}{t} + \frac{\vec{\nabla}p}{\rho} + g\hat{z} &=
            0.\label{eq:nl_mom}.
    \end{align}
\end{subequations}
$\md{}{t} = \pd{}{t} + \p{\vec{u} \cdot \vec{\nabla}}$ is the Lagrangian or
material derivative. $\vec{u}, \rho, p$ denote the velocity field, density and
pressure respectively. We denote $-g\hat{z}$ constant gravitational
acceleration. Note that at hydrostatic equilibrium $\pd{}{t} = 0$ we have
$\vec{\nabla}P_0 = -\rho_0 g\hat{z}$ and so $P_0 = \rho_0 gH$. A shear flow
$u_x(x, z, t) = u_x(z)$ is permitted in hydrostatic equilibrium, but we will
initially consider no shear flow in equilibrium.

@@

\section{Internal Gravity Waves: Theory}\label{s:theory}

@@

In the small perturbation limit, where flow velocities are small compared to the
characteristic space and time scales $\pd{}{t} \gg \vec{u} \cdot \vec{\nabla}$,
we may linearize \autoref{se:nl_upsilon}. We ignore the advective components of the
material derivative and consider small deviations $\Upsilon_1 = \Upsilon -
\Upsilon_0, P_1 = P - P_0$ about hydrostatic equilibrium:
\begin{subequations}\label{se:lin_homo}
    \begin{align}
        \vec{\nabla} \cdot \vec{u} &= 0,\\
        \pd{\Upsilon_1}{t} - \frac{u_{z}}{H} &= 0,\\
        \pd{\vec{u}}{t} + \vec{\nabla}P_1
            + \p{1 + \Upsilon_1} g\hat{z}
            &= 0.
    \end{align}
\end{subequations}

This can be solved to obtain solutions up to undetermined amplitude $A$ the
well-known result \citep{drazin,sutherland0}:
\begin{subequations}\label{se:lin_sol}
    \begin{align}
        u_z\p{x, z, t} &= Ae^{z/2H}e^{i(k_{0x}x + k_{0z}z - \omega_0 t)},\\
        \omega_0^2 &= \frac{N^2k_{0x}^2}{k_{0x}^2 + k_{0z}^2 + \frac{1}{4H^2}},
            \label{eq:disp_rel}
    \end{align}
\end{subequations}
where
\begin{equation}
    N^2 \equiv g^2\p{\rd{\rho}{P} - \at{\pd{\rho}{P}}_{ad}}
        = \frac{g}{H},
\end{equation}
the Brunt-V\"ais\"al\"a frequency is constant

@@

    \item The phase and group velocities can be computed respectively:
        \begin{align}
            \vec{c}_{ph} &=
                \p{\frac{\hat{x}}{k_{0x}} + \frac{\hat{z}}{k_{0z}}}\omega,\\
            \vec{c}_{g} &= \frac{N\p{k_{0x}k_{0z}\hat{x}
                - \p{k_{0z}^2 + \frac{1}{4H^2}}\hat{z}}}
                {\p{k_{0x}^2 + k_{0z}^2 + \frac{1}{4H^2}}^{3/2}} \nonumber\\
                &\approx \frac{\omega k_{0z}}{\abs{\vec{k}_0}^{3/2}}
                    \p{k_{0x}\hat{x} - k_{0z}\hat{z}}.
        \end{align}
        We recover the usual result $\vec{c}_{ph} \cdot \vec{c}_g = 0$
        (e.g.\  \citep{drazin,sutherland1}). Note that an IGW transporting energy
        and momentum upwards $c_{g, z} > 0$ has $k_{0z} < 0$.

    \item The time-averaged total $x$-momentum flux in the $\hat{z}$ direction
        can be computed. Since the linear solution is separable as $f(x, z, t) =
        f(z)e^{i(k_{0x}x - \omega_0 t)}$, $x$ averaging and time averaging are
        equivalent, so we may write
        \begin{align}
            S &\equiv \ev{\rho u_x u_z}_x \equiv
                \frac{1}{L_x}\int_0^{L_x}\limits \rho u_x u_z\;\mathrm{d}x,
                    \label{eq:S_def},\\
                &\approx -\frac{A^2}{2}\rho_0(z=0)\frac{k_{0z}}{k_{0x}},
                    \label{eq:S_lin}
        \end{align}
        where $\ev{\dots}_x$ denotes $x$-averaging.
\end{itemize}

@@

As not to interfere with the incompressibility constraint, we force the system
on the density equation. This constitutes replacing \autoref{eq:nl_up_density}
with
\begin{equation}
    \md{\Upsilon}{t} + u_z\pd{\ln \rho_0}{z}
        = Fe^{-\frac{(z - z_0)^2}{2\sigma^2}}
            \cos \p{k_{0x}x - \omega_0 t}.\label{eq:vol_drive}
\end{equation}
Using a narrow Gaussian profile excites a broad $k_z$ wavenumber spectrum, and
only the $k_{0z}$ satisfying dispersion relation \autoref{eq:disp_rel} for the
given $k_{0x}, \omega_0(k_{0x}, k_{0z})$ will propagate.

In the linearized system, the effect of this forcing can be solved analytically
up to good accuracy: we first approximate the driving term using
$e^{\frac{-(z - z_0)^2}{2\sigma^2}} \approx \sqrt{2\pi \sigma^2}\delta(z -
z_0)$, the $\sigma \to 0$ limit\footnote{In practice, $\sigma$ must be large
enough to be numerically resolved by the spectral code.}. This system is solved
exactly by matching the two homogeneous solutions above and below $z_0$. We may
then approximately relax the solution to nonzero $\sigma$: an extra factor of
$e^{-\frac{(k_{0z}\sigma)^2}{2}}$ arises compared to the $\delta$-function
solution (evaluating the Fourier Transform of $e^{-\frac{(z -
z_0)^2}{2\sigma^2}}$ at $k_z = k_{0z}$), and we obtain
\begin{align}
    u_{1z}&(x, z, t) ={} \frac{Fgk_{0x}^2}{\omega_0^2}
        \frac{1}{2ik_{0z}}\frac{e^{\frac{-(k_{0z}\sigma)^2}{2}}}
        {\sqrt{2\pi\sigma^2}} \times\nonumber\\
        &{}\begin{cases}
        e^{\frac{z - z_0}{2H}}e^{i\p{k_{0x}x + k_{0z}(z - z_0) - \omega_0 t
            + \frac{1}{2k_{0z}H}}}
            & z > z_0\\
        e^{\frac{z - z_0}{2H}}e^{i\p{k_{0x}x - k_{0z}(z - z_0) - \omega_0 t
            + \frac{1}{2k_{0z}H}}}
            & z < z_0\\
    \end{cases}.\label{eq:uz_lin}
\end{align}
The $z > z_0$ solution models an upwards-propagating IGW wavetrain inbound on
the simulation domain from below. The solution for $u_{1x}(x, z, t)$ can be
obtained analytically from \autoref{eq:nl_incomp}.

@@

In developing the results above, we have neglected advective terms $\pd{}{t} \gg
\vec{u} \cdot \vec{\nabla}$. However, since $\abs{\vec{u}} \propto e^{z/2H}$,
any nonzero amplitude IGW will eventually violate the linearity criterion. This
regime is relevant in practice since the IGW may grow to nonlinear amplitudes
before it reaches the surface of the WD \citep{fullerI,fullerII}. We can coarsely
estimate the height at which this happens by when the Lagrangian displacement
$\vec{\xi}$ of a fluid parcel satisfies
\begin{equation}
    \vec{\xi} \cdot \vec{k} \gtrsim 1.\label{eq:nl}
\end{equation}

The general understanding of nonlinear IGW breaking is that the waves'
interactions with the mean flow of the fluid transfer horizontal momentum from
the waves into the fluid; such a process has been conjectured to be responsible
for tidal synchronization in stellar binaries \citep{zahn75,gn89} as well as the
quasi-biennial oscillation \citep{lindzen_qbo}. The details of this process have
been laid out in a few key papers: IGWs are globally unstable to resonant
three-wave interactions \citep{drazin}, causing energy transfer out of the IGW
to daughter modes. These steeper daughter modes facilitate wave breaking,
depositing horizontal momentum in the fluid's mean flow \citep{klostermeyer}.
Once the mean flow reaches the horizontal phase velocity of the parent IGW, a
critical layer forms at which the frequency of the IGW is Doppler-shifted to
zero. The interaction of the parent IGW with the mean flow was first studied in
the inviscid, linear regime in \citep{booker_bretherton}, which found nearly
complete absorption. This result was reproduced with nonzero viscosity
\citep{hazel}, but weakly nonlinear theory \citep{brown_stewartson} and fully
nonlinear simulations \citep{winters1994} suggest that nonlinear effects can
induce reflection.

An estimate of where these mean flow effects become important can be made using
the wave-induced mean flow \citep{eliassen_palm_cite,sutherland0}:
\begin{equation}
     \ev{u_x} = \bar{U}_0 \equiv \frac{1}{L_x}
        \int\limits u_x\;\mathrm{d}x = \frac{\ev{u_xu_z}_x}{c_{g,z}},
        \label{eq:mean_flow}
\end{equation}
where $\ev{\dots}_x$ denotes horizontal averaging. Setting $\bar{U}_0 = c_{ph,
x}$ the critical layer condition, we recover $\frac{u_xk_x}{\omega} = \xi_x k_x
\sim 1$, roughly equivalent to the nonlinearity criterion \autoref{eq:nl}.

\section{Internal Gravity Waves: Linear Numerical Simulation}\label{s:sim}

Towards numerical simulation of IGW breaking, we first verify agreement with
linear theory at weak forcing amplitudes. We perform direct numerical simulation
using the pseudo-spectral code Dedalus \citep{dedalus}. In \autoref{ss:numerics}
we discuss choices of numerical parameters, in \autoref{ss:damping} we
separately discuss use of a damping zone at the top and bottom of the domain,
and in \autoref{ss:lin_ns} we present the results of our ``linear'' simulations,
where we solve the full nonlinear fluid equations for weak forcing.

\subsection{Numerical Setup}\label{ss:numerics}

@@

Denote $L_x, L_z$ to be the physical dimensions of the simulation domain. Our
simulation domain is smaller than our full physical domain, so we adopt
reflection-suppressing boundary conditions. We use periodic boundary conditions
in the $x$ direction and damping layers (described in \autoref{ss:damping}) in
the $z$ direction to damp perturbations exiting the top and bottom of the
domain. We use a Fourier basis in the $x$ direction and either a Chebyshev or
Fourier basis in the $z$ direction; our results were robust to either set of
basis functions. We varied the number of $x, z$ modes (denoted $N_x, N_z$
respectively), and used $3/2$ dealiasing \citep{boyd}.

The time integration uses a split implicit-explicit third-order scheme where
certain terms are treated implicitly and the remaining terms are treated
explicitly. A third-order, four-stage DIRK-ERK scheme \citep{ascher} is used
with adaptive timesteps computed from advective Courant-Friedrichs-Lewy (CFL)
time. Specifically, we use $\Delta t = 0.7 \min(\Delta x / u_x,\Delta z / u_z)$,
where the minimum is taken over every grid point in the domain and $\Delta
x,\Delta z$ are the grid spacings in the $x$ and $z$ directions respectively.

The physics of our simulation is fixed by six parameters: $L_x, L_z$, and
$k_{0x}, \omega_0, F, \nu$. We describe our choices for these parameters below:
\begin{itemize}
    \item $L_z$: Since the wave amplitude grows $\sim e^{z/2H}$, we choose $L_z
        = 10H$ to give $\sim e^4$ apmlitude growth between the damping zones.

@@

    \item $\omega_0$: We choose $\omega_0$ by evaluating dispersion relation
        $\omega_0\p{k_{0x}, k_{0z}}$ for some desired $k_{0z}$.

@@

    \item $F$: We first choose $F$ forcing strength such that $\vec{\xi} \cdot
        \vec{k} \ll 1$ is satisfied everywhere throughout the domain. This
        becomes a constraint on $F$ by using \autoref{eq:uz_lin} and
        $\abs{\vec{u}} \propto e^{z/2H}$.

    \item $\nu$: Nonlinear effects transfer wave energy from $\vec{k}_{0}$ to
        smaller scales. Since well-resolved simulations using spectral methods
        have no inherent numerical viscosity, energy will accumulate at grid
        scales in the absence of artificial dissipation. We introduce
        dissipation parameter $\nu$ used for both artificial viscosity and artificial
        diffusivity. We ensure our equations of motion continue to conserve
        horizontal momentum (see \autoref{ss:strat_impl} for details). Under
        weak forcing, we may set $\nu = 0$, since the energy cascade is
        negligible.
\end{itemize}

@@

To imitate an infinite fluid using a finite simulation domain, we use periodic
boundary conditions in the $x$ direction and damping layers at the top and
bottom of the $z$ direction. This damps waves that reach the edge of the
simulation domain without inducing nonphysical reflection. Specifically, these
are implemented by introducing linear damping terms to \autoref{se:nl_upsilon}:
\begin{align}
    \md{\Upsilon}{t} + u_z \pd{\ln \rho_0}{z} &=
        -\Gamma(z)\Upsilon,\nonumber\\
    \md{\vec{u}}{t} + \vec{\nabla}P + P\vec{\nabla}\Upsilon + g\hat{z} &=
        - \Gamma(z)\vec{u},\nonumber\\
    \frac{1}{2\tau}\s{2 + \tanh \frac{z - z_T}{\Delta z}
        + \tanh \frac{z_B - z}{\Delta z}} &= \Gamma(z),\label{eq:Gamma}
\end{align}
where $z_B = 0.05L_z, z_T = 0.95L_z$ are the boundaries of the damping zone.
This strongly damps perturbations below $z_B$ and above $z_T$ with damping time
$\tau$, negligibly affects dynamics between $z_B, z_T$ and has transition width
governed by $\Delta z$; we use $\Delta z = 0.025L_z$. This prescription, used in
similar studies \citep{lecoanet_damp}, has the advantage of being smooth,
important for spectral methods. The full implementation of the fluid equations
in Dedalus is described in \autoref{ss:strat_impl}.

\subsection{Simulation Results}\label{ss:lin_ns}

We present the results of a simulation satisfying the parameter choices in
\autoref{ss:numerics} such that $\vec{\xi} \cdot \vec{k} \ll 1$ everywhere.
\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{plots/lin_amps.png}
    \caption{Amplitude of excited wave over time in weak forcing simulation,
    in units of the analytical estimate \autoref{eq:uz_lin}. We observe great
    agreement with the linear calculation.}\label{fig:lin_amps}
\end{figure}
\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{plots/lin_fluxes.png}
    \caption{Horizontal momentum flux $S(z, t)$ in units of the analytical estimator
    given by \autoref{eq:S_lin} as a function of $z$ at select times $t$. As
    initial transients die out, the flux converges to be flat below the damping
    layer. The flux excited in the forcing zone is transported without loss to
    the top of the domain, where it is dissipated by the damping layer
    (see \autoref{ss:damping}) without reflection.}\label{fig:lin_fluxes}
\end{figure}

\section{Internal Gravity Waves: Nonlinear Simulation}\label{s:sim}

@@

\section{Acknowledgements}\label{s:ack}
